---
title: "README"
output: md_document
date: "2025-10-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
options(repos = c(CRAN = "https://cloud.r-project.org"))
```

# BayesCOOP - Bayesian Cooperative Learning for Multimodal Integration

The repository houses the **`BayesCOOP`** R package for multimodal integration for prediction of continuous outcomes.

## Dependencies

`BayesCOOP` requires the following `R` package: `devtools` (for installation only). Please install it before installing `BayesCOOP`, which can be done as follows (execute from within a fresh R session):

```{r install, warning=FALSE}
install.packages("devtools")
library(devtools)
```

## Installation

Once the dependencies are installed, `BayesCOOP` can be loaded using the following command:

```{r bcoop, warning=FALSE, message=FALSE}
# # If the package isn't installed, load it from source (package root) for knitting
# if (!requireNamespace("BayesCOOP", quietly = TRUE)) {
#   if (requireNamespace("pkgload", quietly = TRUE)) {
#     # load the current package source without attaching everything to the search path
#     pkgload::load_all(export_all = FALSE, helpers = FALSE, attach_testthat = FALSE)
#   }
# } else {
#   devtools::install_github("himelmallick/BayesCOOP")
#   library(BayesCOOP)
# }
devtools::install_github("himelmallick/BayesCOOP")
library(BayesCOOP)
```

## Example Implementation
## Loading the StelzerDOS real dataset
```{r, message = FALSE, warning=FALSE}
data_train = get(load(url("https://raw.githubusercontent.com/himelmallick/IntegratedLearner/master/data/StelzerDOS.RData"))); rm(pcl)
data_test = get(load(url("https://raw.githubusercontent.com/himelmallick/IntegratedLearner/master/data/StelzerDOS_valid.RData"))); rm(pcl)
```

## Pre-processing the longitudinal data by considering only baseline observations 

### Remove metabolomics from the train set to match with validation
```{r, message = FALSE, warning=FALSE}
if (!requireNamespace("dplyr", quietly = TRUE)) {
  install.packages("dplyr", repos = "https://cloud.r-project.org")
}
library(dplyr)

data_train$feature_metadata = data_train$feature_metadata %>% dplyr::filter(featureType!='Metabolomics')
data_train$feature_table = data_train$feature_table[rownames(data_train$feature_metadata),]
```
 
### Consider only baseline observations for the train set
```{r, message = FALSE, warning=FALSE}
positions = grep("A", colnames(data_train$feature_table), ignore.case = TRUE)
data_train$feature_table = data_train$feature_table[, positions]
data_train$sample_metadata = data_train$sample_metadata[positions, ]
rm(positions)
```

### Consider only baseline observations for the validation set
```{r, message = FALSE, warning=FALSE}
positions = grep("G1", colnames(data_test$feature_table))
data_test$feature_table = data_test$feature_table[, positions]
data_test$sample_metadata = data_test$sample_metadata[positions, ]
rm(positions)
```

```{r BayesCOOP_result, message = FALSE, warning = FALSE, results='hide'}
set.seed(1)
result = BayesCOOP::bayesCoop(data_train, data_test, family = "gaussian", 
                      ss = c(0.05, 1), group = TRUE,
                      bb = TRUE, alpha_dirich = 1, 
                      bbiters = 1100, bbburn = 100, maxit = 100,
                      filter = TRUE, abd_thresh = 0, prev_thresh = 0.1,
                      Warning = TRUE, verbose = TRUE, control = list())

```

```{r}
print(result$mspe)
print(result$time)
```

```{r}
top_indices <- order(abs(result$beta_postmed), decreasing = TRUE)[1:10]
top_values <- result$beta_postmed[top_indices]
if (!requireNamespace("psych", quietly = TRUE)) {
  install.packages("psych", repos = "https://cloud.r-project.org")
}
library(psych)
multi.hist(result$beta_samples[,top_indices],density=TRUE,main="")
```
