---
title: "README"
output: md_document
date: "2025-10-06"
---

# BayesCOOP - Bayesian Cooperative Learning for Multimodal Integration

The repository houses the **`BayesCOOP`** R package for multimodal integration for prediction of continuous outcomes.

## Dependencies

`BayesCOOP` requires the following `R` package: `devtools` (for installation only). Please install it before installing `BayesCOOP`, which can be done as follows (execute from within a fresh R session):

```{r install, eval=FALSE, warning=FALSE}
install.packages("devtools")
library(devtools)
```

## Installation

Once the dependencies are installed, `BayesCOOP` can be loaded using the following command:

```{r bcoop, eval=FALSE, warning=FALSE}
devtools::install_github("himelmallick/BayesCOOP")
library(BayesCOOP)
```

## Example Implementation
## Loading the StelzerDOS real dataset
```{r message = FALSE, warning=FALSE}
data_train = get(load(url("https://raw.githubusercontent.com/himelmallick/IntegratedLearner/master/data/StelzerDOS.RData"))); rm(pcl)
data_test = get(load(url("https://raw.githubusercontent.com/himelmallick/IntegratedLearner/master/data/StelzerDOS_valid.RData"))); rm(pcl)
```

## Pre-processing the longitudinal data by considering only baseline observations 

### Remove metabolomics from the train set to match with validation
```{r message = FALSE, warning=FALSE}
if (!requireNamespace("dplyr", quietly = TRUE)) {
  install.packages("dplyr", repos = "https://cloud.r-project.org")
}
library(dplyr)

data_train$feature_metadata = data_train$feature_metadata %>% dplyr::filter(featureType!='Metabolomics')
data_train$feature_table = data_train$feature_table[rownames(data_train$feature_metadata),]
```
 
### Consider only baseline observations for the train set
```{r message = FALSE, warning=FALSE}
positions = grep("A", colnames(data_train$feature_table), ignore.case = TRUE)
data_train$feature_table = data_train$feature_table[, positions]
data_train$sample_metadata = data_train$sample_metadata[positions, ]
rm(positions)
```

### Consider only baseline observations for the validation set
```{r message = FALSE, warning=FALSE}
positions = grep("G1", colnames(data_test$feature_table))
data_test$feature_table = data_test$feature_table[, positions]
data_test$sample_metadata = data_test$sample_metadata[positions, ]
rm(positions)
```

