---
title: "README"
output: md_document
date: "2025-10-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
options(repos = c(CRAN = "https://cloud.r-project.org"))
```

# BayesCOOP - Bayesian Cooperative Learning for Multimodal Integration

The repository houses the **`BayesCOOP`** R package for multimodal integration for prediction of continuous outcomes.

## Dependencies

`BayesCOOP` requires the following `R` package: `devtools` (for installation only). Please install it before installing `BayesCOOP`, which can be done as follows (execute from within a fresh R session):

```{r install, warning=FALSE, message = FALSE, results = "hide"}
install.packages("devtools")
library(devtools)
```

## Installation

Once the dependencies are installed, `BayesCOOP` can be loaded using the following command:

```{r bcoop, warning=FALSE, message=FALSE, results = "hide"}
devtools::install_github("himelmallick/BayesCOOP", quiet = TRUE)
library(BayesCOOP)
```

## Example Implementation
### Loading the StelzerDOS real dataset
```{r, message = FALSE, warning=FALSE}
data_train = get(load(url("https://raw.githubusercontent.com/himelmallick/IntegratedLearner/master/data/StelzerDOS.RData"))); rm(pcl)
data_test = get(load(url("https://raw.githubusercontent.com/himelmallick/IntegratedLearner/master/data/StelzerDOS_valid.RData"))); rm(pcl)
```

### Pre-processing the longitudinal data by considering only baseline observations 

#### Remove metabolomics from the train set to match with validation
```{r, message = FALSE, warning=FALSE}
if (!requireNamespace("dplyr", quietly = TRUE)) {
  install.packages("dplyr", repos = "https://cloud.r-project.org")
}
library(dplyr)

data_train$feature_metadata = data_train$feature_metadata %>% dplyr::filter(featureType!='Metabolomics')
data_train$feature_table = data_train$feature_table[rownames(data_train$feature_metadata),]
```
 
#### Consider only baseline observations for the train set
```{r, message = FALSE, warning=FALSE}
positions = grep("A", colnames(data_train$feature_table), ignore.case = TRUE)
data_train$feature_table = data_train$feature_table[, positions]
data_train$sample_metadata = data_train$sample_metadata[positions, ]
rm(positions)
```

#### Consider only baseline observations for the validation set
```{r, message = FALSE, warning=FALSE}
positions = grep("G1", colnames(data_test$feature_table))
data_test$feature_table = data_test$feature_table[, positions]
data_test$sample_metadata = data_test$sample_metadata[positions, ]
rm(positions)
```

### Implementing BayesCOOP

```{r BayesCOOP_result, message = TRUE, warning = FALSE, results='hide'}
set.seed(1)
result = BayesCOOP::bayesCoop(data_train, data_test, family = "gaussian", 
                      ss = c(0.05, 1), group = TRUE,
                      bb = TRUE, alpha_dirich = 1, 
                      bbiters = 1100, bbburn = 100, maxit = 100,
                      filter = TRUE, abd_thresh = 0, prev_thresh = 0.1,
                      Warning = TRUE, verbose = TRUE, control = list())

```


### Implementing Cooperative Learning [1]

```{r multiview, message=FALSE}
if (!requireNamespace("multiview", quietly = TRUE)) {
  install.packages("multiview", repos = "https://cloud.r-project.org")
}
library(multiview)

y_train <- BayesCOOP:::gen_datalist(data_train)$y
y_test <- BayesCOOP:::gen_datalist(data_test)$y
xList_train <- BayesCOOP:::gen_datalist(data_train)$xList
xList_test <- BayesCOOP:::gen_datalist(data_test)$xList

## Filtering features
xList_train <- lapply(xList_train, function(foo) BayesCOOP:::filter_features(foo, abd_thresh = 0, prev_thresh = 0.1))
xList_test <- lapply(xList_test, function(foo) BayesCOOP:::filter_features(foo, abd_thresh = 0, prev_thresh = 0.1))

## Considering the common features between train and test set
keep_features <- vector("list", length = length(xList_train))
for(i in 1:length(keep_features)) {
    keep_features[[i]] <- intersect(names(xList_train[[i]]), names(xList_test[[i]]))
    xList_train[[i]] <- as.matrix(xList_train[[i]][, keep_features[[i]], drop = FALSE])
    xList_test[[i]] <- as.matrix(xList_test[[i]][, keep_features[[i]], drop = FALSE])
}

xList_train <- lapply(xList_train, function(foo) as.matrix(foo))
xList_test <- lapply(xList_test, function(foo) as.matrix(foo)) ## need matrix form for multiview

# Function to perform cross-validation for choosing alpha for cooperative learning application
cvfit_alpha <- function(xlist, y, rho, alpha, fold) {
  cvfit <- cv.multiview(x_list = xlist, y = y, rho = rho,
                       alpha = alpha, nfolds = fold)
  lmin <- cvfit$lambda.min
  lmin_indx <- which(cvfit$lambda == lmin)
  cverr_min <- cvfit$cvm[lmin_indx]
  return(list(lmin = lmin, lmin_indx = lmin_indx, cverr_min = cverr_min,
              alpha = alpha))
}

start.time <- Sys.time()
cv_results_early <- cvfit_alpha(xlist = xList_train, y = y_train, rho = 0, alpha = 1, fold = 5)
opt_lambda_early <- cv_results_early$lmin
fit_multiview_early <- multiview(xList_train, y_train, lambda = opt_lambda_early, rho = 0, alpha = 1)
stop.time <- Sys.time()
time_multiview_early <- as.numeric(round(difftime(stop.time, start.time, units="min"), 3), units = "mins")
rm(start.time); rm(stop.time)
beta_hat_early <- as.vector(fit_multiview_early$beta)
## prediction
y_pred_early <- predict(fit_multiview_early, newx = xList_test, s = opt_lambda_early, rho = 0, alpha = 1)
## mean squared prediction error 
mspe_multiview_early <- mean((y_pred_early - y_test) ^ 2)

start.time <- Sys.time()
cv_results_intermediate <- cvfit_alpha(xlist = xList_train, y = y_train, rho = 0.5, alpha = 1, fold = 5)
opt_lambda_intermediate <- cv_results_intermediate$lmin
fit_multiview_intermediate <- multiview(xList_train, y_train, lambda = opt_lambda_intermediate, rho = 0.5, alpha = 1)
stop.time <- Sys.time()
time_multiview_intermediate <- as.numeric(round(difftime(stop.time, start.time, units="min"), 3), units = "mins")
rm(start.time); rm(stop.time)
beta_hat_intermediate <- as.vector(fit_multiview_intermediate$beta)
## prediction
y_pred_intermediate <- predict(fit_multiview_intermediate, newx = xList_test, s = opt_lambda_intermediate, rho = 0.5, alpha = 1)
## mean squared prediction error 
mspe_multiview_intermediate <- mean((y_pred_intermediate - y_test) ^ 2)

start.time <- Sys.time()
cv_results_late <- cvfit_alpha(xlist = xList_train, y = y_train, rho = 1, alpha = 1, fold = 5)
opt_lambda_late <- cv_results_late$lmin
fit_multiview_late <- multiview(xList_train, y_train, lambda = opt_lambda_late, rho = 1, alpha = 1)
stop.time <- Sys.time()
time_multiview_late <- as.numeric(round(difftime(stop.time, start.time, units="min"), 3), units = "mins")
rm(start.time); rm(stop.time)
beta_hat_late <- as.vector(fit_multiview_late$beta)
## prediction
y_pred_late <- predict(fit_multiview_late, newx = xList_test, s = opt_lambda_late, rho = 1, alpha = 1)
## mean squared prediction error 
mspe_multiview_late <- mean((y_pred_late - y_test) ^ 2)
```

### Comparing BayesCOOP with Cooperative Learning

```{r predictive accuracy comparison, message=FALSE}
## BayesCOOP
print(result$mspe)
print(result$time)

## Early Fusion
print(mspe_multiview_early)
print(time_multiview_early)
## Intermediate Fusion
print(mspe_multiview_intermediate)
print(time_multiview_intermediate)
## Late Fusion
print(mspe_multiview_late)
print(time_multiview_late)
```


## References

[1] Daisy Yi Ding, Shuangning Li, Balasubramanian Narasimhan, and Robert Tibshirani (2022).
“Cooperative Learning for Multiview Analysis.” Proceedings of the National Academy of Sciences (USA), Vol. 119, No. 38, e2202113119. DOI: [10.1073/pnas.2202113119](https://www.pnas.org/doi/full/10.1073/pnas.2202113119)
