---
title: "README"
output: md_document
date: "2025-08-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# BayesCOOP

The repository contains the analysis codes from the Bayesian coopertive project.

## Dependencies

`BayesCOOP` requires the following `R` package: `devtools` (for installation only). Please install it before installing `BayesCOOP`, which can be done as follows (execute from within a fresh R session):

```r
install.packages("devtools")
library(devtools)
```
## Installation
 
Once the dependencies are installed, `BayesCOOP` can be loaded using the following command:
```r
devtools::install_github("himelmallick/BayesCOOP")
library(BayesCOOP)
```

## Load libraries 
```{r message = FALSE, warning=FALSE}
library(BhGLM)
library(tidyverse)
library(caret)
```

## sourcing required functions
```{r message = FALSE, warning=FALSE}
source("https://github.com/himelmallick/BayesCOOP/blob/master/R/helperFunctions.R")
```

## Loading the StelzerDOS real dataset
```{r message = FALSE, warning=FALSE}
data_train = get(load(url("https://raw.githubusercontent.com/himelmallick/IntegratedLearner/master/data/StelzerDOS.RData"))); rm(pcl)
data_test = get(load(url("https://raw.githubusercontent.com/himelmallick/IntegratedLearner/master/data/StelzerDOS_valid.RData"))); rm(pcl)
```

## Pre-processing the longitudinal data by considering only baseline observations 

### Remove metabolomics from the train set to match with validation
```{r message = FALSE, warning=FALSE}
data_train$feature_metadata = data_train$feature_metadata %>% dplyr::filter(featureType!='Metabolomics')
data_train$feature_table = data_train$feature_table[rownames(data_train$feature_metadata),]
```

### Consider only baseline observations for the train set
```{r message = FALSE, warning=FALSE}
positions = grep("A", colnames(data_train$feature_table), ignore.case = TRUE)
data_train$feature_table = data_train$feature_table[, positions]
data_train$sample_metadata = data_train$sample_metadata[positions, ]
rm(positions)
```

### Consider only baseline observations for the validation set
```{r message = FALSE, warning=FALSE}
positions = grep("G1", colnames(data_test$feature_table))
data_test$feature_table = data_test$feature_table[, positions]
data_test$sample_metadata = data_test$sample_metadata[positions, ]
rm(positions)
```

# BayesCoop Implementation 
```r 
set.seed(1)
result = bayesCoop(data_train, data_test, family = "gaussian", 
                   ss = c(0.05, 1), group = TRUE,
                   bb = TRUE, alpha_dirich = 1, 
                   bbiters = 1100, bbburn = 100, maxit = 100,
                   filter = TRUE, abd_thresh = 0, prev_thresh = 0.1,
                   Warning = TRUE, verbose = TRUE)
```

```{r BayesCOOP_result, message = FALSE, warning = FALSE, echo = FALSE, results='hide'}
set.seed(1)
result = bayesCoop(data_train, data_test, family = "gaussian", 
                   ss = c(0.05, 1), group = TRUE,
                   bb = TRUE, alpha_dirich = 1, 
                   bbiters = 1100, bbburn = 100, maxit = 100,
                   filter = TRUE, abd_thresh = 0, prev_thresh = 0.1,
                   Warning = TRUE, verbose = TRUE)
```


```{r}
result$mspe
result$time
```


Let's check MCMC convergence of the reciprocal Bayesian LASSO estimator through two visualizations: trace plots and histograms.

```{r message = FALSE, warning = FALSE}
######################################
# Visualization of Posterior Samples #
######################################

##############
# Trace Plot #
##############

top_indices <- order(result$beta_postmed, decreasing = TRUE)[1:10]
top_values <- result$beta_postmed[top_indices]


library(coda)
par(mar=c(2,2,2,2))
plot(mcmc(result$beta_samples[,top_indices]),density=FALSE,smooth=TRUE)
```

```{r message = FALSE, warning = FALSE}
#############
# Histogram #
#############

library(psych)
multi.hist(result$beta_samples[,top_indices],density=TRUE,main="")
```
