---
title: "BayesCOOP: End-to-End Tutorial"
author: "Saptarshi Roy, Sreya Sarkar, Himel Mallick"
output:
  html_document:
    theme: readable
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: false
date: "2025-10-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
options(repos = c(CRAN = "https://cloud.r-project.org"))
set.seed(1)
```

# Overview

**BayesCOOP** is a scalable Bayesian framework for supervised multimodal data integration. It currently supports continuous outcomes.

BayesCOOP combines fast maximum a posteriori (MAP) estimation with Bayesian bootstrap–based uncertainty quantification. Using a *jittered* group spike-and-slab LASSO shrinkage prior, it jointly selects informative modalities and features and produces accurate predictions with principled uncertainty estimates while remaining computationally efficient.

In this tutorial, we will:

1.  Load and preprocess a longitudinal pregnancy dataset with multiple omics layers (proteomics + immune profiling).
2.  Fit BayesCOOP.
3.  Fit a Cooperative Learning baseline ([Ding et al., 2022](https://www.pnas.org/doi/10.1073/pnas.2202113119)).
4.  Compare predictive performance and runtime.

We will refer to the pregnancy cohort data from [Stelzer et al. (2021)](<https://www.science.org/doi/10.1126/scitranslmed.abd9898>), which profiled longitudinal immune and proteomic changes in late gestation. We treat this as a supervised prediction problem (continuous outcome such as gestational age or days to spontaneous labor).

We will also compare against Cooperative Learning ([Ding et al., 2022](https://www.pnas.org/doi/10.1073/pnas.2202113119)), which is a widely used multiview baseline.

------------------------------------------------------------------------

# 1. Installation and Package Setup

You can install the development version of BayesCOOP from GitHub:

```{r install_bayescoop, eval=FALSE}
install.packages("devtools")
devtools::install_github("himelmallick/BayesCOOP")
library(BayesCOOP)
```

We will also need `dplyr` for data manipulation and `multiview` for Cooperative Learning (Ding et al.).

```{r packages}
if (!requireNamespace("dplyr", quietly = TRUE)) {
  install.packages("dplyr")
}
if (!requireNamespace("multiview", quietly = TRUE)) {
  install.packages("multiview")
}

library(dplyr)
library(multiview)
library(BayesCOOP)
```

------------------------------------------------------------------------

# 2. Load the Example Dataset

```{r load_data}
data_train <- get(load(url(
  "https://raw.githubusercontent.com/himelmallick/IntegratedLearner/master/data/StelzerDOS.RData"
)))
data_test <- get(load(url(
  "https://raw.githubusercontent.com/himelmallick/IntegratedLearner/master/data/StelzerDOS_valid.RData"
)))
```

------------------------------------------------------------------------

# 3. Preprocessing

```{r preprocess}
data_train$feature_metadata <- data_train$feature_metadata %>%
  dplyr::filter(featureType != "Metabolomics")

data_train$feature_table <- data_train$feature_table[
  rownames(data_train$feature_metadata), , drop = FALSE
]

pos_train <- grep("A", colnames(data_train$feature_table), ignore.case = TRUE)
data_train$feature_table  <- data_train$feature_table[, pos_train, drop = FALSE]
data_train$sample_metadata <- data_train$sample_metadata[pos_train, , drop = FALSE]

pos_test <- grep("G1", colnames(data_test$feature_table))
data_test$feature_table   <- data_test$feature_table[, pos_test, drop = FALSE]
data_test$sample_metadata <- data_test$sample_metadata[pos_test, , drop = FALSE]
```

------------------------------------------------------------------------

# 4. Fit BayesCOOP

```{r run_bayescoop}
fit_bayescoop <- BayesCOOP::bayesCoop(
  data_train,
  data_test,
  family       = "gaussian",
  ss           = c(0.05, 1),
  group        = TRUE,
  bb           = TRUE,
  alpha_dirich = 1,
  bbiters      = 1100,
  bbburn       = 100,
  maxit        = 100,
  filter       = TRUE,
  abd_thresh   = 0,
  prev_thresh  = 0.1,
  Warning      = TRUE,
  verbose      = TRUE,
  control      = list()
)

fit_bayescoop$mspe
fit_bayescoop$rho_postmed
fit_bayescoop$time
```

------------------------------------------------------------------------

# 5. Cooperative Learning Baseline (Ding et al.)

```{r coop_baseline}
y_train <- BayesCOOP:::gen_datalist(data_train)$y
y_test  <- BayesCOOP:::gen_datalist(data_test)$y
xList_train <- BayesCOOP:::gen_datalist(data_train)$xList
xList_test  <- BayesCOOP:::gen_datalist(data_test)$xList

xList_train <- lapply(
  xList_train,
  function(foo) BayesCOOP:::filter_features(foo, abd_thresh = 0, prev_thresh = 0.1)
)
xList_test <- lapply(
  xList_test,
  function(foo) BayesCOOP:::filter_features(foo, abd_thresh = 0, prev_thresh = 0.1)
)

for (i in seq_along(xList_train)) {
  keep_i <- intersect(names(xList_train[[i]]), names(xList_test[[i]]))
  xList_train[[i]] <- as.matrix(xList_train[[i]][, keep_i, drop = FALSE])
  xList_test[[i]]  <- as.matrix(xList_test[[i]][, keep_i, drop = FALSE])
}

cvfit_alpha <- function(xlist, y, rho, alpha = 1, folds = 5) {
  cvfit <- multiview::cv.multiview(x_list = xlist, y = y, rho = rho, alpha = alpha, nfolds = folds)
  list(lambda_min = cvfit$lambda.min)
}

rhos_to_try <- c("Early" = 0, "Intermediate" = 0.5, "Late" = 1)
coop_results <- lapply(names(rhos_to_try), function(label) {
  rho_val <- rhos_to_try[[label]]
  start_time <- Sys.time()
  cvres <- cvfit_alpha(xList_train, y_train, rho = rho_val)
  best_lambda <- cvres$lambda_min
  fit_mv <- multiview::multiview(xList_train, y_train, lambda = best_lambda, rho = rho_val, alpha = 1)
  y_pred_mv <- predict(fit_mv, newx = xList_test, s = best_lambda, rho = rho_val, alpha = 1)
  mspe_mv <- mean((y_pred_mv - y_test)^2)
  runtime_min <- round(as.numeric(difftime(Sys.time(), start_time, units = "mins")), 3)
  list(method = label, rho = rho_val, mspe = mspe_mv, time_min = runtime_min)
})

coop_results
```

------------------------------------------------------------------------

# 6. Performance Comparison

```{r compare_performance}
bayescoop_summary <- list(
  method    = "BayesCOOP (Bayesian bootstrap)",
  mspe      = fit_bayescoop$mspe,
  time_min  = fit_bayescoop$time
)

perf_table <- rbind(
  data.frame(method = bayescoop_summary$method, mspe = bayescoop_summary$mspe, time_min = bayescoop_summary$time_min),
  do.call(rbind, lapply(coop_results, function(res) {
    data.frame(method = paste0("Cooperative Learning (", res$method, ", rho=", res$rho, ")"),
               mspe = res$mspe, time_min = res$time_min)
  }))
)

perf_table
```

------------------------------------------------------------------------

# 7. Reproducibility

```{r sessioninfo}
sessionInfo()
```

------------------------------------------------------------------------
